The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-02-09 article/article.tex
asymmetric witness trees

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\begin{document}

@@

Condsider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\lceil\frac{n}{d}\rceil$ \todo{simplify to n/d?}. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymetric tie-breaking} in combination with \emph{partitioning}\todo{check} reduces the bound for the gap to $\Theta \left( \frac{\log\log n}{d  \cdot \log \Phi_d}\right)$ \todo{$\Phi_d$ is undefined}.

@@

The \emph{Always-go-left} algorithm works as follows. Partition the $n$ bins into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\lceil\frac{n}{d}\rceil$ . We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, put the ball into the bin with smallest index $i$.

@@

@@

\subsection{Greedy}
\label{sec:analysis:greedy}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The same ball may be represented by several nodes. The nodes and edges of the witness tree represent events that may occur or not. For simplicity, we assume that all events represented by a witness tree are stochastically independent.

In the following, we donate the $d$ bins that are selected for a ball as the \emph{locations} of a ball. That is, before inserting a ball $d$ locations are generated and the ball is then inserted into one of those. We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the $i$th child of $u$. Then $e$ represents the event that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}

@@

\subsubsection{Witness tree construction}
\label{sec:constructionSymWT}
We construct a witness tree for the event that some bin contains $L+4$ balls as follows:
\begin{compactenum}
\item The topmost ball $b_1$ in bin $x$ is assigned to the root node $v_r$. Note that $b_1$ is the last ball that was inserted into the bin $x$.
\item Since the Greedy-scheme puts ball $b_1$ into bin $x$ only if each of the $d$ sampled locations contain at least the same number of balls at bin $x$, we can conclude that each of the $d$ locations contain at least $L+3$ balls at the time ball $b_1$ was inserted. Thus, the $d$ children of the root node are assigned the topmost balls of these $d$ bins.
\item Consider these child nodes and continue the assignment of the previous step recursively until the leaf nodes are reached.  
\end{compactenum}

@@

\subsubsection{Probability of activation}
\label{sec:probabilitySymWT}
In this section, we derive the probability that some witness tree of order $L$ is activated, which gives an upper bound for the probability that the maximum load is $L+3$ at some time $t$. The probability $p_a$ that some witness tree of order $L$ is activated is bounded by the number of witness trees $w$ times the upper bound on the probability $p_1$ that a specific witness tree is activated.

@@

Thus, the probability that an activated symmetric witness tree exists at time $t$ is at most 
\begin{align}
p_{swt} &= n^{k} \cdot \left(\frac{d}{n}\right)^{k-1} \cdot 3^{-d \cdot l} \\
		&= n \cdot d^{k-1} \cdot 3^{-d \cdot l}
\end{align}

Using $k \leq 2l$ we get
\begin{align}
p_{swt} &\leq n \cdot d^{2\cdot l} \cdot 3^{-d \cdot l} \\
		&= n \cdot \left(d^2\right)^l \cdot \left( 3^{-d} \right)^l
\end{align}
which we simplify by using $2d^2 \leq 3^d$
\begin{align}
p_{swt} &\leq n \cdot \left(\frac{3^d}{2}\right)^l \cdot \left( 3^{-d}\right)^l \\
		&= n \cdot 2^{-l} \cdot 3^{d\cdot l} \cdot 3^{-d \cdot l} \\
		&= n\cdot 2^{-d^L}		
\end{align}

Finally, we can show that the activation of a symmetric witness tree with distinct balls is unlikely, if the order of the tree $L$ is large enough. For any constant $\alpha > 0 $, let
\begin{align}
L \geq \log_d\log_2 n + \log_d\left(1+\alpha\right)
\end{align}
Then, 
\begin{align}
p_{swt} &\leq n \cdot 2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n \cdot 2^{-{d}^{\log_d\log_2 n}\cdot d^{\log_d \cdot\left(1+\alpha\right)}} \\
		&= n \cdot \left( 2^{-\log_2 n \cdot \left(1+ \alpha \right)} \right) \\
		&= n \cdot n^{-\left(1+\alpha\right)} \\
		&= n^{-\alpha}
\end{align}

\subsection{Always-go-left}
\label{sec:analysis:alg}
Similarly to Section \ref{sec:analysis:greedy}, where we used symmetric witness trees to upper-bound the probability that a bin contains too many balls when using the \emph{Greedy} scheme, we will use asymmetric witness trees to get an analogous upper bound for the \emph{Always-go-left} scheme.

With one exception, the definition of an \emph{asymmetric witness tree} is the same as the definition of a symmetric witness tree, including the definition of activation. Only its structure differs from that of a symmetric tree. An asymmetric witness tree has the same structure as a $d$-ary Fibonacci tree $T_d\left(k \right)$, which is defined as follows.
\begin{compactitem}
\item $T_d(1)$ is a single node
\item $T_d(2) = T_d(1)$  
\item For $3\leq k \leq d$, $T_d(k)$ is a tree with $k-1$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(1)$.
\item For $k>d$, $T_d(k)$ is a tree with $d$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(k-d)$.
\end{compactitem}
\todo{illustrate}

@@

\subsubsection{Witness tree construction}
\label{sec:analysis:constructionAsymWT}

\subsubsection{Probability of activation}
\label{sec:analysis:probabilityAsymWT}

>>>>

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{invariant}{Invariant}
\begin{document}

@@

Condsider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ \todo{simplify to n/d?}. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymetric tie-breaking} in combination with \emph{partitioning}\todo{check} reduces the bound for the gap to $\Theta \left( \frac{\log\log n}{d  \cdot \log \Phi_d}\right)$ \todo{$\Phi_d$ is undefined}.

@@

The \emph{Always-go-left} algorithm works as follows. Partition the $n$ bins into $d$ groups, each of size $\left\lfloor \frac{n}{d}\right\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ . We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, put the ball into the bin with smallest index $i$.

@@

Throughout this section, we donate the $d$ bins that are selected for a ball as the \emph{locations} of a ball. That is, before inserting a ball $d$ locations are generated and the ball is then inserted into one of those.

@@

\subsection{Greedy}
\label{sec:analysis:greedy}
\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The same ball may be represented by several nodes. The nodes and edges of the witness tree represent events that may occur or not. For simplicity, we assume that all events represented by a witness tree are stochastically independent.

We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the $i$th child of $u$. Then $e$ represents the event that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}

@@

\subsubsection{Construction}
\label{sec:analysis:constructionSymWT}
We construct a witness tree for the event that some bin contains $L+4$ balls as follows:
\begin{compactenum}
\item The topmost ball $b_1$ in bin $x$ is assigned to the root node $v_r$. Note that $b_1$ is the last ball that was inserted into the bin $x$.
\item Since the Greedy-scheme puts ball $b_1$ into bin $x$ only if each of the $d$ sampled locations contain at least the same number of balls at bin $x$, we can conclude that each of the $d$ locations contain at least $L+3$ balls at the time ball $b_1$ was inserted. Thus, the $d$ children of the root node are assigned the topmost balls of these $d$ bins.
\item Consider these child nodes and continue the assignment of the previous step recursively until the leaf nodes are reached.  
\end{compactenum}

@@

\subsubsection{Probability of activation}
\label{sec:analysis:probabilitySymWT}
In this section, we derive the probability that some witness tree of order $L$ is activated, which gives an upper bound for the probability that the maximum load is $L+3$ at some time $t$. The probability $p_a$ that some witness tree of order $L$ is activated is bounded by the number of witness trees $w$ times the upper bound on the probability $p_1$ that a specific witness tree is activated.

@@

Thus, the probability that an activated symmetric witness tree exists at time $t$ is at most 
\begin{align}
p_{swt} &= w \cdot p_e \cdot p_l \\
		&= n^{k} \cdot \left(\frac{d}{n}\right)^{k-1} \cdot 3^{-d \cdot l} \\
		&= n \cdot d^{k-1} \cdot 3^{-d \cdot l}
\end{align}

Using $k \leq 2l$ we get
\begin{align}
p_{swt} &\leq n \cdot d^{2\cdot l} \cdot 3^{-d \cdot l} \\
		&= n \cdot \left(d^2\right)^l \cdot \left( 3^{-d} \right)^l
\end{align}
which we simplify by using $2d^2 \leq 3^d$
\begin{align}
p_{swt} &\leq n \cdot \left(\frac{3^d}{2}\right)^l \cdot \left( 3^{-d}\right)^l \\
		&= n \cdot 2^{-l} \cdot 3^{d\cdot l} \cdot 3^{-d \cdot l} \\
		&= n\cdot 2^{-l}		\\
		&= n\cdot 2^{-d^L}		
\end{align}

For any constant $\alpha > 0 $, let
\begin{align}
L \geq \log_d\log_2 n + \log_d\left(1+\alpha\right)
\end{align}
and $n$ be large enough such that 
\begin{align}
L \in \Theta \left(\ln \ln n \right)
\end{align}
Then, 
\begin{align}
p_{swt} &\leq n \cdot 2^{{-d}^{\log_d\log_2 n + \log_d\left(1+\alpha\right)}} \\
		&= n \cdot 2^{-{d}^{\log_d\log_2 n}\cdot d^{\log_d \cdot\left(1+\alpha\right)}} \\
		&= n \cdot \left( 2^{-\log_2 n \cdot \left(1+ \alpha \right)} \right) \\
		&= n \cdot n^{-\left(1+\alpha\right)} \\
		&= n^{-\alpha}
\end{align}

\todo{add the asymptotic bound}

\subsection{Always-go-left}
\label{sec:analysis:alg}
Similarly to Section \ref{sec:analysis:greedy}, where we used symmetric witness trees to upper-bound the probability that a bin contains too many balls when using the \emph{Greedy} scheme, we will use asymmetric witness trees to get an analogous upper bound for the \emph{Always-go-left} scheme. We will make the same simplification as stated in Assumption \ref{assumption:independence}, namely that all the events represented by the edges and nodes of a witness tree are stochastically independent.

\subsubsection{Asymmetric witness trees}
\label{sec:analysis:definitionAsymWT}
With one exception, the definition of an \emph{asymmetric witness tree} is the same as the definition of a symmetric witness tree, including the definition of activation. Only its structure differs from that of a symmetric tree. An asymmetric witness tree has the structure of a $d$-ary Fibonacci tree $T_d\left(k \right)$, which is defined as follows.
\begin{compactitem}
\item $T_d(1)$ is a single node.
\item $T_d(2) = T_d(1)$.
\item For $3\leq k \leq d$, $T_d(k)$ is a tree with $k-1$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(1)$.
\item For $k>d$, $T_d(k)$ is a tree with $d$ children, which are the roots of the trees $T_d(k-1),\ldots,T_d(k-d)$.
\end{compactitem}
\todo{illustrate}

@@

For each node we define a label $(h, i)$, where $h, i \in \mathbb{N}$, $0 \leq h \leq L$ and  $1\leq i \leq d$. 
\begin{compactitem}
\item A node that is the root of a Fibonacci tree $T_d(1)$ is assigned the label $(0,1)$.
\item Analogously, a node that is the root of a Fibonacci tree $T_d(2)$ is assigned the label $(0,2)$. Thus, nodes labelled $(0,1)$ or $(0,2)$ do not have children.
\item A node with label $(0, i)$, where $i>2$, has $i-1$ children with labels $(0,i-1)\dots (0,1)$
\item A node with label $(h,i)$ , where $h>0$, has $d$ children with the labels $(h,i-1)\dots(h,1)$,$(h-1,d)\dots(h-1, i)$.
\item The root of a Fibonacci tree $T_d(d\cdot h+i)$ has the label $(h, i)$. 
\end{compactitem}
Thus, our the root node of our witness tree will be labelled $(L, 1)$. Note that the labels are unique within the set of children of a node $v$, but not for all nodes of the tree.

\subsubsection{Construction}
\label{sec:analysis:constructionAsymWT}
The labels defined in the previous section are useful to state the following invariant, which will be maintained while assigning balls to the nodes of the witness tree.

\begin{invariant}
Ball $b_v$ referenced by a node $v$ with label $(h,i)$ was placed in a bin belonging to group $i$. The bin contained at least $h+3$ other balls at the time of insertion of the ball $b_v$.
\end{invariant}

Let $x$ be the bin that contains $L+4$ balls. The root node gets assigned the topmost ball in bin $x$. Each of the $d$ locations of this ball points to a bin with at least $L+3$ balls in it, at the time of its insertion. These are the $d$ bins that were sampled by the Always-go-left scheme and thus bin $i$ belongs to the $i$'th group of bins, where each group consists of $n/d$ bins\footnote{For the sake of simplicity, we assume that $d$ divides $n$ evenly, i.e., that all groups have the exact same size. It can be shown that the analysis still holds if each group is of size $\lfloor\frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$.}. The child node of the root with label $(L-1, i)$ is assigned the topmost ball of the bin the $i$' location points to. The assignment of balls to nodes than proceeds recursively. Given a node $v$ with label $(h, i)$ whose ball $b_v$ was put into some bin belonging to group $i$ and containing at least $l+3$ other balls, we can conclude the following.
\begin{compactitem}
\item For $1\leq j < i$, the $j$'th location of the ball $b_v$ points to a bin containing at least $l+4$ balls at the insertion time of the ball $b_v$. If the $j$'th location contained less balls than the $i$'th location, the Always-go-left scheme would have put the ball $b_v$ into the bin pointed to by location $j$. The topmost ball in location $j$ is assigned to the  child $u$ of $v$ with label $(l,j)$. 
\item For $i < j \leq d$, the $j$'th location contains at least $l+3$ balls. If location $j$ contained less, the Always-go-left scheme would have put the ball $b_v$ into the bin it points to. The topmost ball in the bin that location $j$ points to is assigned to the child node $u$ of $v$ labelled with $(l-1, j)$. \todo{the paper claims that the locations $j$ point to bins with $l+2$ balls. This can not be true, since the topmost, i.e., the $l+2$ ball - call it $b_t$ - gets assigned to a node with label $(l'=l-1, j)$. Since there are $l+1=l'+2$ balls below it, at the time of the placement of ball $b_t$, the bin $j$ it gets put into has only $l'+2$ balls in it, violating the invariant. Same with the case $j<i$.} 
\end{compactitem}

\subsubsection{Probability of activation}
\label{sec:analysis:probabilityAsymWT}
Analogously to the symmetric witness tree, the possible number of asymmetric witness trees is upper-bounded by $w=n^k$ and the probability that all edge events occur is\footnote{As in Section \ref{sec:analysis:greedy}, $k$ denotes the number of nodes of the witness tree and $l$ the number of leaves.} 
\begin{align}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align}

The probability that all leaf events occur too is the same as for the symmetric witness tree, but its derivation differs. Again, at any time at most $b'=n/3$ bins contain 3 or more balls, but we do not know how those bins are distributed among the groups. Let $b_i$ be the number of bins in group $i$ that contain 3 or more balls. Then $\beta_i = b_i\cdot d/ n$ is the fraction of the bins $b_i$ in group $i$. Since no more than $n/3$ bins can contain 3 or more balls, the values $b_i$ are subject to the constraint
\begin{align}
\frac{n}{d}\sum_{i=1}^d \beta_i \leq \frac{n}{3}
\end{align}
and thus 
\begin{align}
\sum_{i=1}^d \beta_i \leq \frac{d}{3}
\end{align}

The probability that all $d$ locations of a ball point to a bin containing 3 or more balls is then 
\begin{align}
p_{l_1} = \prod_{i=1}^{d} \beta_i
\end{align}
In other words, a leaf event occurs with probability $p_{l_1}$, which is maximized when all $b_i$ are equal. We thus set $b_i \leq 1/3$ for all $i$ and get as upper bound for the probability that all leaf events occur 
\begin{align}
p_l \leq 3^{-d}
\end{align}

The above results imply the same probability for the activation of an asymmetric witness tree as we derived in Section \ref{sec:analysis:probabilitySymWT} for the activation of a symmetric witness tree, that is
\begin{align}
p_{awt} &\leq n 2^{-l}
\end{align}
However, the number of leaves is larger, since 
\begin{align}
l \geq \Phi_d^{d\cdot L -1}
\end{align}
holds\footnote{see Section \ref{sec:analysis:definitionAsymWT} for details}.

For any constant $\alpha > 0$, let 
\begin{align}
L = \left\lceil{\frac{\ln\log_2 n + \ln\left(1+\alpha\right)}{d\cdot \ln \Phi_d}}\right  \rceil +1
\end{align}
and $n$ be large enough such that
\begin{align}
L \in \Theta\left(\frac{\ln\ln n}{d\cdot \ln \Phi_d}\right)
\end{align}
